---
title: Simulation exercise in<br />**Genomik und Bioinformatik**
output: html_document
---

```{r simulation_options, include=FALSE}
library(knitr)
opts_chunk$set(fig.width=6, fig.height=6, fig.path='RmdFigs/', dpi=72,
               warning=FALSE, message=FALSE, prompt=TRUE, cache=FALSE)
set.seed(53079237)
```

Estimating $\pi$
================

$\pi$ is a number that can be used to compute the area of a
circle. The unit circle is the circle with radius 1. In order to
approximate $\pi$, we can generate uniformly distributed random points
in a square of length and height 1. By counting the number of points
which are within the unit circle and dividing this number by all
points generated, we can approximate $\pi/4$.

### 1. R script to estimate $\pi$

Write a script in R that
approximates $\pi$ as described above. Randomly generate 1000 points
in the unit square and estimate $\pi$.

**Answer:** $\pi$ can be estimated as follows:

```{r estimate_pi}
num.samples <- 1000
rand_points <- matrix(runif(2 * num.samples), ncol=2)
dist <- sqrt((rand_points[, 1])^2 + (rand_points[, 2])^2)
in.circle <- ifelse(dist <= 1, TRUE, FALSE)
pi.estimate <- 4 * sum(in.circle) / num.samples
```


**Question (a):** Is `runif` or `rnorm` the better choice to generate
uniformly distributed random points in the unit square? Why?

**Answer:** `rnorm` generates random numbers according to a normal
distribution. I.e., many points related to normal destributed random
numbers are close to the origin of the unit square, while only a few
such points are far away. Hence, this is not the proper distribution
for uniformly distributed points in the unit square. On the other
hand, `runif` with the two parameters 0 and 1 generates uniformly
distributed random numbers between 0 and 1. Points that are formed of
two such numbers, are uniformly distributed in the unit square.


**Question (b):** For illustration, generate a plot that shows
random points within the unit circle in red and random points
outside the unit circle in blue. Why is this plot useful?

**Answer:** The plot allows visual control of the script and
illustrates how $\pi$ is estimated:

```{r estimate_pi_plot, fig.cap="Estimate pi in a unit square by simulation.", fig.align="center"}
cols <- ifelse(in.circle, "red", "blue")
plot(rand_points[, 1], rand_points[, 2], col=cols, xlab="x", ylab="y", pch=20, xaxs="i", yaxs="i",
     main="Points in unit square\n(red = in unit circle, blue = out of unit circle)")
x <- seq(0, 1, 0.001)
lines(x, sqrt(1 - x^2), lwd=2)
```
     

**Question (c):** How close is the estimation to the correct value, if you use
1000 random points? Is the result the same, if you re-run your
script? Explain the observed behavior.

**Answer:** The estimation is pretty close to the correct value 
(estimated: `r pi.estimate`, correct: `r pi`, absolute difference: 
`r pi.estimate - pi`, relative difference: `r pi.estimate / pi`).
In this case, re-running the .Rmd script yields the same results because the
seed of the random number generator is set to a fixed value at the beginning of
the .Rmd script. If the seed was not set at the beginning the result
would differ.

----

### 2. Precision improvement by increasing the number of samples

Go through the tasks below to approximate $\pi$ 100 times using 100 
samples, 100 times using 1000 samples and 100 times using 10,000 samples.

**Question (a):** Write a function to approximate $\pi$ as described above. The 
function should take the number of points as argument and return the 
estimate.

**Answer:** The function to describe $\pi$ is as follows:
```{r estimate_pi2}
myApprox <- function(N=1000){
  x <- runif(N)   
  y <- runif(N)
  circle <- y <= sqrt(1 - x**2)   
  counts <- length(which(circle))    
  p <-(counts/N)*4   
  
  return(p)
}
myApprox()
```

**Question (b)**: Create a 100 x 3 matrix to store the estimates from the single
runs of the function.

**Answer:** Preallocate the required space for the matrix:
```{r estimate_pi3}
myMatrix <- matrix(0, nrow  = 100, ncol = 3)
```


**Question (c)**: Use the `for` construct to generate two nested loops
in R. The inner loop is used to re-run $\pi$-estimation 100 times
using the same parameter (e.g. number of samples per
estimation). The outer loop is used to perform the inner loop
using different number of samples. Fill the matrix with the estimates.

**Answer:** Fill the matrix:
```{r estimate_pi4}
samples <- c(100, 1000, 10000)  
for (n in seq_along(samples)) {     
  for (m in 1:100) {  
    myMatrix[m, n] <- myApprox(samples[n])  
  }
}
```

**Question (d)**: Use the function `summary` to report the results for different 
sample sizes. What is shown in the output of `summary` under 
`'1st Qu'` and `'3rd Qu'`?

```{r estimate_pi5}
summary(myMatrix)
```
**Answer:** Explaination to `summary()`:
Data is divided into four quantiles in total. If 1st quantile is 3.04 that means that 25% of values (for that column) are lower than 3.04. Same is true for 3rd quantile, which equals to the 75% lowest values. The mean is the "2nd quantile" so to say.


**Question (e)**: The function `summary` shows averages and extremes
etc.  of a distribution. Alternatively, A distribution can be
illustrated by a histogram.  Use the function `hist` to visualize the
distributions of $\pi$-estimations for different sample sizes.  How
can you use the plotting parameter `mfrow` to show the histograms for
different parameters one above the other in the same plot? (Hint:
`help(par)`) Use the same range on the x axis for every histogram to
make the visual comparison of the plots easier.  Use a reasonable
number of breaks in the histograms. Mark the true value of $\pi$ with
a vertical red line in every histogram and use `mtext` to add labels
to these lines.

**Answer:** Following the instructions above we get the histograms with the following code:
```{r estimate_pi6}
N <- c(100, 1000, 10000)
myDimensions <- dim(myMatrix)
myColumns <- myDimensions[2]

par(mfrow=c(myColumns,1))
for (j in seq(1, myColumns)){
  hist(myMatrix[, j], main=paste((N[j]), "samples"), xlim=c(2.9,3.4))
  abline(v=mean(myMatrix[, j]), col="blue")
  abline(v=pi, col="red")
  mtext("pi", side = 1, col = "red", line = 1)
}
```

Session Information
-------------------

The following versions of R and R packages were used to generate the report
above:

```{r sessionInfo, echo=FALSE}
sessionInfo()
```
